---
title: "Covid Hospitalization and Vaccination Study"
author: "Gregory Maynard"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

# Required Packages and Global Variables

required packages:

```{r load-packages, message=FALSE}
library(openintro)

library(tidyverse)
library(lubridate)
library(sf)

# for map data
library(albersusa)

# for SODA API calls
library(httr)
library(jsonlite)
```


Global Vars:
 
```{r}

# hospital and vaccination data endpoints to soda api
hospital_data_soda_endpoint <- "https://healthdata.gov/resource/anag-cw7u.json"
vaccination_data_soda_endpoint <- "https://data.cdc.gov/resource/8xkx-amqh.json"

# needed to avoid soda api throttling
soda_app_key <- "3tl81UiFuoaJ1DboeqMuy5K9U"

# api url for arcgis hrr boundary shape data
hrr_boundary_arcgis_endpoint <- "https://services.arcgis.com/nzS0F0zdNLvs7nc8/arcgis/rest/services/HRR_boundaries/FeatureServer"

# url for hrr zip crosswalk zip file from Dartmouth Atlas
crosswalk_hrr_zip_url <- "https://data.dartmouthatlas.org/downloads/geography/ZipHsaHrr19.csv.zip"

#cached data
cached_data_folder_base = "cached-data"
cached_bed_data_folder = paste(cached_data_folder_base, "hospital_bed_usage", sep = "/")
cached_vaccination_data_folder = paste(cached_data_folder_base, "vaccination_rates", sep = "/")
cached_zip_hrr_crosswalk_folder = paste(cached_data_folder_base, "crosswalk", sep = "/")

cached_hrr_shapes_filepath = paste(cached_data_folder_base, "hhr_shapes_arcgis_data.shape", sep = "/")
cached_hospital_bed_data_filepath = paste(cached_bed_data_folder, "hospital_bed_usage_healthdatagov_", sep = "/")
cached_vaccination_data_filepath = paste(cached_vaccination_data_folder, "vaccination_rates_cdc_", sep = "/")

cached_zip_hrr_crosswalk_compressed_filepath = paste(cached_zip_hrr_crosswalk_folder, "hrr_zip_crosswalk.zip", sep = "/")
cached_zip_hrr_crosswalk_filepath = paste(cached_zip_hrr_crosswalk_folder, "ZipHsaHrr19.csv", sep = "/")

```


# Helper Functions

```{r}
# Returns a dataframe from the given soda api endpoint. 
get_soda_dataframe <- function(endpoint, simple_filter = NULL, where_filter, select_filter, limit = 1000){
  
  #Build the query
  soda_query = paste0(endpoint, 
            "?",                                      
             simple_filter,
             "&$where=", where_filter,
             "&$select=", select_filter,
             "&$limit=", limit,
             "&$$app_token=", soda_app_key)
  
  print(paste("Query: ", soda_query))
  
  # Getting response details in API
  response_details = GET(url = URLencode(soda_query))
  
  # Getting status of HTTP Call
  response_status_code = status_code(response_details)
  
  print(paste0("SODA http response: ", response_status_code))
  
  # Converting content to text
  response_content_text = content(response_details,
                            "text", encoding = "UTF-8")
   
  # Parse Json Data
  response_content_json = fromJSON(response_content_text,
                             flatten = TRUE)
   
  # Converting into dataframe
  soda_dataframe = as.data.frame(response_content_json)
}

######### HOSPITAL BED DATA#########

# gets dataframe with total inpatient beds, used inpatient beds, and used inpatient beds for covid patients, 
# for each hospital on record for the week given (date). This request asks for the first 10,000 records. A typical
# week on this dataset has between 4,000 and 5,000 records.
#
# columns selected from dataset: 
#               state
#               hospital_name
#               city
#               fips_code
#               inpatient_beds_7_day_avg
#               inpatient_beds_used_7_day_avg
#               inpatient_beds_used_covid_7_day_avg
get_bed_utilization_healthdata_gov_soda_api <- function(date){
  
  get_soda_dataframe(
    endpoint = hospital_data_soda_endpoint,
    #simple_filter = "state='WA'",                              # can add this for smaller dataset for testing (fast soda api call)
    where_filter = paste0("collection_week='",date, "'"),
    select_filter = "state,hospital_name,city,fips_code,inpatient_beds_7_day_avg,inpatient_beds_used_7_day_avg,inpatient_beds_used_covid_7_day_avg",
    limit = 7000) %>% 
    
    # Converting value types #
    
    # convert values to numeric
    mutate(
      fips_code = as.numeric(fips_code),
      inpatient_beds_7_day_avg = as.numeric(inpatient_beds_7_day_avg),
      inpatient_beds_used_7_day_avg = as.numeric(inpatient_beds_used_7_day_avg),
      inpatient_beds_used_covid_7_day_avg = as.numeric(inpatient_beds_used_covid_7_day_avg)) %>% 

    # replace negative bed counts with NA
    mutate(
      inpatient_beds_7_day_avg = ifelse(inpatient_beds_7_day_avg > -1, inpatient_beds_7_day_avg, NA),
      inpatient_beds_used_7_day_avg = ifelse(inpatient_beds_used_7_day_avg > -1, inpatient_beds_used_7_day_avg, NA),
      inpatient_beds_used_covid_7_day_avg = ifelse(inpatient_beds_used_covid_7_day_avg > -1, inpatient_beds_used_covid_7_day_avg, NA))
}

# Gets hospital bed usage dataframe.
# Checks for data in the cached data folder first, then looks for the data 
# using the soda rest api. If no data file is found locally, the downloaded 
# file is saved in the cached data folder.
#
# columns in dataset: 
#               state
#               hospital_name
#               city
#               fips_code
#               inpatient_beds_7_day_avg
#               inpatient_beds_used_7_day_avg
#               inpatient_beds_used_covid_7_day_avg
get_bed_utilization_data <- function(date){
  
  # this database likes this format
  date <- ymd(date)
  
  # first we need to determine the filename of the cached data file.
  filepath = paste0(cached_hospital_bed_data_filepath, date, ".csv")
  
  if( !file.exists(filepath) ) {
    
    # get from healthdata.gov
    
    print("Hospital bed cached data was not found. Retrieving from healthdata.gov (soda api)...")
    bed_data = get_bed_utilization_healthdata_gov_soda_api(date = date)
    
    write_csv(bed_data, filepath)
    print("File retrieved and cached for future use.")
  }
  else {
    
    # get from cached data file
    
    bed_data = read_csv(filepath, show_col_types = FALSE)
    print("Hospital bed data loaded from local cache.")
  }
  
  return(bed_data)
}


######### VACCINATION DATA #########

# gets dataframe with vaccination records from the CDC using the soda api
# converts the column's data types to those loaded by read_csv by default 
# so that data returned from the soda endpoint is the same as the data loaded 
# from the cache file.
#
# There are around 3300 records per date in this dataset.
#
# columns selected from dataset: 
#               fips
#               series_complete_pop_pct
#               administered_dose1_pop_pct
#               booster_doses_vax_pct
get_vaccination_rates_cdc_soda_api <- function(date){
  
  data = get_soda_dataframe(
    endpoint = vaccination_data_soda_endpoint,
#    simple_filter = "recip_state='WA'",                              # can add this for smaller dataset for testing (fast soda api call)
    where_filter = paste0("date='",date, "'"),
    select_filter = "date,fips,series_complete_pop_pct,administered_dose1_pop_pct,booster_doses_vax_pct,completeness_pct,census2019",
    limit = 10000)
  
## Data needs to be converted so it is consistent. We want it to load from the csv 
## the same way it is retrieved from the soda endpoint
  
  data <- data %>% mutate(
    date = as_date(date))

  if("fips" %in% colnames(data))
    data <- data %>% mutate (fips = as.numeric(fips))
  
  if("series_complete_pop_pct" %in% colnames(data))
    data <- data %>% mutate (series_complete_pop_pct = as.numeric(series_complete_pop_pct))

  if("administered_dose1_pop_pct" %in% colnames(data))
    data <- data %>% mutate (administered_dose1_pop_pct = as.numeric(administered_dose1_pop_pct))
  
  if("booster_doses_vax_pct" %in% colnames(data))
    data <- data %>% mutate (booster_doses_vax_pct = as.numeric(booster_doses_vax_pct))

  return(data)
}


# gets dataframe with vaccination records.
# first checks for a locally cached data file
# If no data file is found locally, the downloaded file
# is saved in the cached data folder.
#
# columns selected from dataset: 
#               fips
#               series_complete_pop_pct
#               administered_dose1_pop_pct
#               booster_doses_vax_pct
get_vaccination_rates_data <- function(date){
  
  # ymd is how arguments should be entered and files should be saved
  date <- ymd(date)
  
  # first we need to determine the path/name of the cached data file.
  filepath = paste0(cached_vaccination_data_filepath, date, ".csv")
  
  if( !file.exists(filepath) ) {
    # get from CDC
    
    print("Vaccination Rates cached data was not found. Retrieving from data.cdc.gov (soda api)...")
    data = get_vaccination_rates_cdc_soda_api(date = date)
    
    write_csv(data, filepath)
    print("Vaccination records file retrieved and cached for future use.")
  }
  else {
    # get from cached data file
    
    data = read_csv(filepath, show_col_types = FALSE)
    print("Vaccination records loaded from local cache.")
  }
  
  return(data)
}


######### HRR BOUNDARY MAP DATA ######### 

# get hrr shapefile data from arcgis
get_hrr_shapes_arcgis <- function() {
  
  if( !file.exists(cached_hrr_shapes_filepath) ) {
  
    print("No cached hrr shapefile was found. Downloading from arcgis and saving to file...")
    
    # get and save the file
    url = parse_url(hrr_boundary_arcgis_endpoint)
    url$path = paste(url$path, "0/query", sep = "/")
    url$query = list(where = "HRRNUM > -1",
                      outFields = "*",
                      returnGeometry = "true",
                      f = "geojson")
    request = build_url(url)
    
    hrr_shape_data = st_read(request)
    
    st_write(hrr_shape_data, cached_hrr_shapes_filepath, driver = "GeoJSON")
  }
  else {
    hrr_shape_data = st_read(cached_hrr_shapes_filepath)
    print("HRR shapes data loaded from local cache.")
  }
  
  return(hrr_shape_data)
}

# Get hrr zipcode crosswalk from Dartmouth Atlas zip file
get_hrr_zip_crosswalk <- function(){
  if( !file.exists(cached_zip_hrr_crosswalk_compressed_filepath) ) {
    print("Downloadeing Zip HRR Crosswalk data from Dartmouth Atlas.")
    download.file(crosswalk_hrr_zip_url, cached_zip_hrr_crosswalk_compressed_filepath)
    print("Unzipping...")
    unzip(cached_zip_hrr_crosswalk_compressed_filepath, exdir = cached_zip_hrr_crosswalk_folder)  # unzip your file 
  }
  
  print("Loading crosswalk data from file.")
  return(read.csv(cached_zip_hrr_crosswalk_filepath) %>% select(zipcode19, hrrnum))
}


#### misc ####

create_data_folder_if_dne <- function(){
  if ( !dir.exists(file.path(cached_data_folder_base)) ) {
    print("Creating cached data folder...")
    dir.create(cached_data_folder_base)
    dir.create(cached_bed_data_folder)
    dir.create(cached_vaccination_data_folder)
    dir.create(cached_zip_hrr_crosswalk_folder)
  }
  else
    print("Founds cached data folder.")
}

# ggplot theme function
my_map_theme <- function(){
  theme(panel.background=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        axis.title=element_blank())
}

```



# Data Sources Explained

## TL/DR

Data sources:

1. Hospital Bed Usage in USA - per hospital
2. Vaccination Rates in USA - per county
3. HRR Geography in USA - per HRR number
4. County Geography in USA - per county
5. Crosswalk for Zip Code and HRR number

Anytime a new file is downloaded, that file is cached in the "cached-data" folder. Anytime a request is made 
for a dataset by date, this folder is checked first.

## Local Data Caching

Vaccination data, bed data, hrr boundary data, and ZIP Code HRR Crosswalks are obtained by downloaded 
from the internet. Only the needed portions of the datasets are requested from the corresponding endpoints. 
Before downloading, this application first checks if the dataset has already been downloaded in a local cache 
file. Whenever a new portion of a dataset is downloaded, it is saved to the cache folder local to the application 
folder.

Setup the data caching folder:

```{r}
create_data_folder_if_dne()
```

## Vaccination Rates Data per US County

Vaccination rates are obtained from the CDC: https://data.cdc.gov/Vaccinations/COVID-19-Vaccinations-in-the-United-States-County/8xkx-amqh. 
This dataset is large, so instead of downloading the whole dataset, this application accesses it through the SODA API and retrieves only the 
relevant rows and columns.

The "COVID-19 Vaccinations in the United States,County" data provides counts and percentages of people who have been 
vaccinated in each county of the United States.

The variables retrieved are:

* **fips**
* **series_complete_pop_pct**: "Percent of people who have completed a primary series (have second dose of a two-dose vaccine or one dose of a single-dose vaccine) based on the jurisdiction and county where vaccine recipient lives."
* **administered_dose1_pop_pct**: "Percent of Total Pop with at least one Dose by State of Residence"
* **booster_doses_vax_pct**: "Percent of people who completed a primary series and have received a booster (or additional) dose."

## Hospital Capacity Data for USA per Hospital

Hospital bed usage counts is obtained from HealthData.gov: https://healthdata.gov/Hospital/COVID-19-Reported-Patient-Impact-and-Hospital-Capa/anag-cw7u. 
This dataset is large, so instead of downloading the whole dataset, this application accesses it through the SODA API and retrieves only the 
relevant rows and columns.

The "COVID-19 Reported Patient Impact and Hospital Capacity by Facility" data provides counts on hospital bed utilization that is aggregated weekly.

The variables retrieved are:

* **Hospital_name** and **fips_code**
* **inpatient_beds_7_day_avg**: "Average number of total number of staffed inpatient beds in your hospital including all overflow, observation, and active surge/expansion beds used for inpatients (including all ICU beds) reported in the 7-day period."
* **inpatient_beds_used_7_day_avg**: "Average of total number of staffed inpatient beds that are occupied reported during the 7-day period."
* **inpatient_beds_used_covid_7_day_avg**: "Average of reported patients currently hospitalized in an inpatient bed who have suspected or confirmed COVID-19 reported during the 7-day period."

*Inpatient bed counts are used instead of total bed counts because many hospitals that only have inpatient bed data do not include data for inpatient and outpatient totals.*


#### Calculate ratios for hospital bed usage

We also calculate: 

* **bed_usage_ratio**: The ratio of hospital beds used out of 100 beds (inpatient)
* **covid_bed_usage_ratio**: The ratio of hospital beds used by covid patients out of 100 beds (inpatient)
* **covid_bed_usage_total_bed_usage_ratio**: The ratio of hospital beds used by covid patients out of all used beds (inpatient)

#### Calculate average ratios per region

The hospital bed dataset contains records for each hospital. For mapping, each map region will need to represent the average 
of all hospitals present in that region.


## Geographic Shape Data

### US County Shape Data

County shape data is obtained from the Albersusa package.

Load county data using the Alber's projection:

```{r}
# get county shapes for mapping
us_county_shape_data <- counties_sf("laea") %>% 
  mutate(fips = as.numeric(as.character(fips)))
```

### Hospital Referral Region Shape Data

Shape data for USA HRR regions are downloaded from arcgis.com using their "FeatureServer" REST api.

https://www.arcgis.com/home/item.html?id=46bf6790c4e0455e9379ee9769b1a5ab

HRR number to zip code translation (2019)

A Crosswalk to correlate zip codes with hrr numbers is needed. This crosswalk is obtained from Dartmouth Atlas as a zip file.
this file is unzipped and loaded. 

https://data.dartmouthatlas.org/supplemental/#crosswalks

```{r}
# Get hrr shapes for mapping
hrr_shape_data <- get_hrr_shapes_arcgis()
zip_hrr_crosswalk_data <- get_hrr_zip_crosswalk()
```

# GGPLOT Maps

### Graph by Date Functions

```{r}
### Date must be given in yyyymmdd format
Graph_Vaccination_Rates_By_County <- function(date) {
  vaccination_data = get_vaccination_rates_data(date = date)
  
  vaccination_data = us_county_shape_data %>% 
    left_join(vaccination_data, by = "fips") 
    
  vaccination_data %>% 
    ggplot() +
    geom_sf(aes(fill = series_complete_pop_pct/100)) +
    scale_fill_continuous("Fully Vaccinated", low="yellow", high="red", labels = scales::percent) +
    ggtitle("COVID Vaccination Status", subtitle = "Percentage of county population that is fully vaccinated (from CDC)") +
    my_map_theme()
}

### Date must be given in yyyymmdd format
Graph_Hospital_Bed_Usage_By_County <- function(date){
  hospital_bed_data <- get_bed_utilization_data(date = date)
  
  # calculate ratios
  hospital_bed_data <- hospital_bed_data %>% 
    mutate(
      bed_usage_ratio = inpatient_beds_used_7_day_avg / inpatient_beds_7_day_avg * 100,
      covid_bed_usage_ratio = inpatient_beds_used_covid_7_day_avg / inpatient_beds_7_day_avg * 100,
      covid_bed_usage_total_bed_usage_ratio = inpatient_beds_used_covid_7_day_avg / inpatient_beds_used_7_day_avg
    )
  
  
  county_bed_ratios <- hospital_bed_data %>% 
    group_by(fips_code) %>% 
    summarise(
      bed_usage_ratio = mean(bed_usage_ratio, na.rm = TRUE), 
      covid_bed_usage_ratio = mean(covid_bed_usage_ratio, na.rm = TRUE), 
      covid_bed_usage_total_bed_usage_ratio = mean(covid_bed_usage_total_bed_usage_ratio, na.rm = TRUE))
  
  
  
  us_county_shape_data %>% 
    left_join(county_bed_ratios, by = c("fips" = "fips_code")) %>% 
    
    ggplot() +
      geom_sf(aes(fill = covid_bed_usage_ratio/100)) +
      scale_fill_continuous("Beds Used for Covid", low="deepskyblue", high="green", labels = scales::percent) +
      ggtitle("Hospital Bed Usage", subtitle = "Percentage of county's total hospital beds used by covid patients") +
      my_map_theme()
}

```


### Vaccination Rates Map

```{r}
Graph_Vaccination_Rates_By_County("2021/03/1")

Graph_Vaccination_Rates_By_County("2021/09/24")

```



## Hospital Bed Usage Map

#### By County

```{r}
Graph_Hospital_Bed_Usage_By_County("2020/11/06")

Graph_Hospital_Bed_Usage_By_County("2022/07/22")
```

#### By HRR





...

