---
title: "Covid Hospitalization and Vaccination Study"
author: "Gregory Maynard"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(openintro)

library(tidyverse)

# for map data
library(albersusa)

# for SODA API calls
library(httr)
library(jsonlite)
```


# Variables

```{r}
# hospital data endpoint to soda api
hospital_data_soda_endpoint <- "https://healthdata.gov/resource/anag-cw7u.json"

# needed to avoid api throttling
healthdata_soda_app_token <- "3tl81UiFuoaJ1DboeqMuy5K9U"
```


# Helper Functions

```{r}
# Returns a dataframe from the given soda api endpoint.
get_soda_dataframe <- function(endpoint, simple_filter = NULL, where_filter, select_filter, limit = 1000, app_token){
  
  #Build the query
  soda_query = paste0(hospital_data_soda_endpoint, 
            "?",                                      
             simple_filter,
             "&$where=", where_filter,
             "&$select=", select_filter,
             "&$limit=", limit,
             "&$$app_token=", app_token)
  
  # Getting response details in API
  response_details = GET(url = URLencode(soda_query))
  
  # Getting status of HTTP Call
  response_status_code = status_code(response_details)
  
  print(paste0("SODA http response: ", response_status_code))
  
  # Converting content to text
  response_content_text = content(response_details,
                            "text", encoding = "UTF-8")
   
  # Parse Json Data
  response_content_json = fromJSON(response_content_text,
                             flatten = TRUE)
   
  # Converting into dataframe
  soda_dataframe = as.data.frame(response_content_json)
}

# gets dataframe with total inpatient beds, used inpatient beds, and used inpatient beds for covid patients, 
# for each hospital on record for the week given (date)
get_healthdata_gov_bed_utilization_from_soda <- function(date){
  get_soda_dataframe(
    endpoint = hospital_data_soda_endpoint,
    
  # simple_filter = "state='WA'",  # can add for smaller dataset for testing (fast soda api call)
    
    where_filter = "collection_week='2023-02-24'",
    select_filter = "state,hospital_name,city,fips_code,inpatient_beds_7_day_avg,inpatient_beds_used_7_day_avg,inpatient_beds_used_covid_7_day_avg",
    limit = 10000,
    app_token = healthdata_soda_app_token) %>% 
    
    # convert values to numeric
    mutate(
      fips_code = as.numeric(fips_code),
      inpatient_beds_7_day_avg = as.numeric(inpatient_beds_7_day_avg),
      inpatient_beds_used_7_day_avg = as.numeric(inpatient_beds_used_7_day_avg),
      inpatient_beds_used_covid_7_day_avg = as.numeric(inpatient_beds_used_covid_7_day_avg)) %>% 

    # replace negative bed counts with NA
    mutate(
      inpatient_beds_7_day_avg = ifelse(inpatient_beds_7_day_avg > -1, inpatient_beds_7_day_avg, NA),
      inpatient_beds_used_7_day_avg = ifelse(inpatient_beds_used_7_day_avg > -1, inpatient_beds_used_7_day_avg, NA),
      inpatient_beds_used_covid_7_day_avg = ifelse(inpatient_beds_used_covid_7_day_avg > -1, inpatient_beds_used_covid_7_day_avg, NA))
  

}

# ggplot theme function
my_map_theme <- function(){
  theme(panel.background=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        axis.title=element_blank())
}

```


# Data Sources

## Hospital Capacity Data

Hospital bed usage counts is obtained from HealthData.gov: https://healthdata.gov/Hospital/COVID-19-Reported-Patient-Impact-and-Hospital-Capa/anag-cw7u. This dataset is over 300 MB, so instead of downloading the whole dataset, this application accesses it through the SODA API.

The "COVID-19 Reported Patient Impact and Hospital Capacity by Facility" provides data on hospital bed utilization that is aggregated weekly.

The variables we are using are:
* **Hospital_name** and **fips_code**
* **inpatient_beds_7_day_avg**: Average number of total number of staffed inpatient beds in your hospital including all overflow, observation, and active surge/expansion beds used for inpatients (including all ICU beds) reported in the 7-day period.
* **inpatient_beds_used_7_day_avg**: Average of total number of staffed inpatient beds that are occupied reported during the 7-day period.
* **inpatient_beds_used_covid_7_day_avg**: Average of reported patients currently hospitalized in an inpatient bed who have suspected or confirmed COVID-19 reported during the 7-day period.

*Inpatient bed counts are used instead of total bed counts because many hospitals that only have inpatient bed data do not include data for inpatient and outpatient totals.*

We also calculate: 
* **bed_usage_ratio**: The ratio of hospital beds used out of 100 beds (inpatient)
* **covid_bed_usage_ratio**: The ratio of hospital beds used by covid patients out of 100 beds (inpatient)
* **covid_bed_usage_total_bed_usage_ratio**: The ratio of hospital beds used by covid patients out of all used beds (inpatient)

```{r}

# get data
hospital_bed_data <- get_healthdata_gov_bed_utilization_from_soda(date = '2023-02-24')

# calculate ratios
hospital_bed_data <- hospital_bed_data %>% 
  mutate(
    bed_usage_ratio = inpatient_beds_used_7_day_avg / inpatient_beds_7_day_avg * 100,
    covid_bed_usage_ratio = inpatient_beds_used_covid_7_day_avg / inpatient_beds_7_day_avg * 100,
    covid_bed_usage_total_bed_usage_ratio = inpatient_beds_used_covid_7_day_avg / inpatient_beds_used_7_day_avg
  )

```


Get Map Geometry

```{r}

us_county_map_geometry <- counties_sf("laea") %>% 
  mutate(fips = as.numeric(as.character(fips)))

wa_counties_map_geometry <- counties_sf() %>% 
  mutate(fips = as.numeric(as.character(fips))) %>% 
  filter(iso_3166_2 == "WA")


```

Hospitals are grouped by county using their fips code and then averages for bed ratios are calculated for each county.

```{r}
bed_ratios <- hospital_bed_data %>% 
  group_by(fips_code) %>% 
  summarise(
    bed_usage_ratio = mean(bed_usage_ratio, na.rm = TRUE), 
    covid_bed_usage_ratio = mean(covid_bed_usage_ratio, na.rm = TRUE), 
    covid_bed_usage_total_bed_usage_ratio = mean(covid_bed_usage_total_bed_usage_ratio, na.rm = TRUE))

```



graph covid bed usage ratio

```{r}

wa_counties_map_geometry %>% 
  left_join(bed_ratios, by = c("fips" = "fips_code")) %>% 
  ggplot() +
  geom_sf(aes(fill = covid_bed_usage_ratio)) +
  my_map_theme()

us_county_map_geometry %>% 
  left_join(bed_ratios, by = c("fips" = "fips_code")) %>% 
  ggplot() +
  geom_sf(aes(fill = covid_bed_usage_ratio)) +
  my_map_theme()

```


...

